---
title: "Ocean Rescue Accessibility Across Oʻahu"
author: "DJ"
---

# Introduction

Ocean rescue accessibility is an important part of emergency preparedness on Oʻahu, Hawaiʻi.

This project examines three types of coastal emergency accessibility:

1.  Swimming and shoreline rescue access
2.  Lifeguard response
3.  Coast Guard marine response

# Research Question

**How does estimated accessibility to swimming rescue, lifeguard response, and Coast Guard services vary across Oʻahu?**

# Data Sources

- [Honolulu Lifeguard Towers](https://data.honolulu.gov/d/8crx-qhct)
- [Hawaiʻi GIS Body Surfing Sites](https://geodata.hawaii.gov/arcgis/rest/services/CoastalMarine/MapServer/9)
- [Hawaiʻi GIS Shoreline Access](https://geodata.hawaii.gov/arcgis/rest/services/CoastalMarine/MapServer/25)
- [Hawaiʻi GIS Data Catalog](https://planning.hawaii.gov/gis/download-gis-data-expanded/)
- [USCG Sector Honolulu](https://www.pacificarea.uscg.mil/Our-Organization/Oceania-District/Oceania-District-Units/Sector-Honolulu/Sector-Honolulu-Units/)
- [NOAA Electronic Navigational Charts](https://charts.noaa.gov/ENCs/ENCs.shtml)

```{r}
#| label: setup

library(sf)
library(dplyr)
library(leaflet)
library(ggplot2)
library(knitr)
```

```{r}
#| label: load-gis-data

data_dir <- "data"

required_files <- c(
  "lifeguards.geojson",
  "body_surfing.geojson",
  "shoreline_access.geojson",
  "oahu_roads.geojson"
)

missing_files <- required_files[
  !file.exists(file.path(data_dir, required_files))
]

if (length(missing_files) > 0) {
  stop(
    "Missing files: ",
    paste(missing_files, collapse = ", ")
  )
}

lifeguards <- st_read(
  file.path(data_dir, "lifeguards.geojson"),
  quiet = TRUE
) |> st_transform(4326)

body_surfing <- st_read(
  file.path(data_dir, "body_surfing.geojson"),
  quiet = TRUE
) |> st_transform(4326)

shoreline <- st_read(
  file.path(data_dir, "shoreline_access.geojson"),
  quiet = TRUE
) |> st_transform(4326)

roads <- st_read(
  file.path(data_dir, "oahu_roads.geojson"),
  quiet = TRUE
) |> st_transform(4326)

# Coast Guard facility table is optional.
coast_guard_file <- file.path(
  data_dir,
  "coast_guard_facilities.csv"
)

if (file.exists(coast_guard_file)) {
  coast_guard <- read.csv(coast_guard_file)
} else {
  coast_guard <- data.frame(
    name = character(),
    address = character()
  )
}
```

```{r}
#| label: filter-oahu

oahu_bbox <- st_bbox(
  c(
    xmin = -158.30,
    ymin = 21.20,
    xmax = -157.60,
    ymax = 21.75
  ),
  crs = st_crs(4326)
)

oahu_extent <- st_as_sfc(oahu_bbox)

lifeguards <- st_filter(
  lifeguards, oahu_extent
)

body_surfing <- st_filter(
  body_surfing, oahu_extent
)

shoreline <- st_filter(
  shoreline, oahu_extent
)

roads <- st_filter(
  roads, oahu_extent
)
```

```{r}
#| label: dataset-summary

summary_table <- data.frame(
  Dataset = c(
    "Lifeguard Towers",
    "Body Surfing Sites",
    "Shoreline Access Points",
    "Road Segments"
  ),
  Features = c(
    nrow(lifeguards),
    nrow(body_surfing),
    nrow(shoreline),
    nrow(roads)
  )
)

knitr::kable(
  summary_table,
  caption = "GIS features within the preliminary Oʻahu extent"
)
```

```{r}
#| label: overview-map

overview_map <- leaflet(
  options = leafletOptions(preferCanvas = TRUE)
) |>
  addProviderTiles("CartoDB.Positron") |>
  addPolylines(
    data = roads,
    color = "gray",
    weight = 1,
    opacity = 0.25,
    group = "Road Network"
  ) |>
  addCircleMarkers(
    data = lifeguards,
    radius = 6,
    color = "red",
    group = "Lifeguard Towers"
  ) |>
  addCircleMarkers(
    data = body_surfing,
    radius = 5,
    color = "blue",
    group = "Swimming Sites"
  ) |>
  addCircleMarkers(
    data = shoreline,
    radius = 4,
    color = "orange",
    group = "Shoreline Access"
  ) |>
  addLayersControl(
    overlayGroups = c(
      "Road Network",
      "Lifeguard Towers",
      "Swimming Sites",
      "Shoreline Access"
    ),
    options = layersControlOptions(
      collapsed = FALSE
    )
  ) |>
  setView(
    lng = -157.95,
    lat = 21.48,
    zoom = 10
  )

overview_map
```

# Lifeguard Response Accessibility

```{r}
#| label: lifeguard-calculations

lifeguards_utm <- st_transform(
  lifeguards, 32604
)

surf_utm <- st_transform(
  body_surfing, 32604
)

stopifnot(
  nrow(lifeguards_utm) > 0,
  nrow(surf_utm) > 0
)

nearest_lifeguard <- st_nearest_feature(
  surf_utm,
  lifeguards_utm
)

lifeguard_distance <- st_distance(
  surf_utm,
  lifeguards_utm[nearest_lifeguard, ],
  by_element = TRUE
)

surf_utm$lifeguard_distance_km <-
  as.numeric(lifeguard_distance) / 1000

# Illustrative speed, not observed response speed.
lifeguard_speed_kmh <- 5

surf_utm$lifeguard_time_min <-
  surf_utm$lifeguard_distance_km /
  lifeguard_speed_kmh * 60

surf_utm$lifeguard_category <- cut(
  surf_utm$lifeguard_time_min,
  breaks = c(-Inf, 5, 10, 20, Inf),
  labels = c(
    "0-5 min",
    "5-10 min",
    "10-20 min",
    "20+ min"
  )
)
```

```{r}
#| label: lifeguard-table

lifeguard_results <- surf_utm |>
  st_drop_geometry() |>
  select(
    lifeguard_distance_km,
    lifeguard_time_min,
    lifeguard_category
  )

knitr::kable(
  head(lifeguard_results, 15),
  digits = 2,
  caption = "Illustrative lifeguard proximity results"
)
```

```{r}
#| label: lifeguard-map

time_colors <- c(
  "0-5 min" = "#2ECC71",
  "5-10 min" = "#F1C40F",
  "10-20 min" = "#E67E22",
  "20+ min" = "#E74C3C"
)

lifeguard_pal <- colorFactor(
  palette = time_colors,
  domain = levels(surf_utm$lifeguard_category)
)

lifeguard_map <- leaflet() |>
  addProviderTiles("CartoDB.Positron") |>
  addCircleMarkers(
    data = st_transform(surf_utm, 4326),
    radius = 7,
    color = ~lifeguard_pal(lifeguard_category),
    fillColor = ~lifeguard_pal(lifeguard_category),
    fillOpacity = 0.85,
    popup = ~paste0(
      "Nearest tower: ",
      round(lifeguard_distance_km, 2),
      " km<br>",
      "Hypothetical time: ",
      round(lifeguard_time_min, 1),
      " min"
    )
  ) |>
  addCircleMarkers(
    data = lifeguards,
    radius = 5,
    color = "blue",
    group = "Lifeguard Towers"
  ) |>
  addLegend(
    position = "bottomright",
    pal = lifeguard_pal,
    values = surf_utm$lifeguard_category,
    title = "Hypothetical Time"
  ) |>
  setView(
    lng = -157.95,
    lat = 21.48,
    zoom = 10
  )

lifeguard_map
```

# Swimming and Shoreline Rescue Accessibility

```{r}
#| label: swimming-calculations

shore_utm <- st_transform(
  shoreline, 32604
)

stopifnot(nrow(shore_utm) > 0)

nearest_shore <- st_nearest_feature(
  surf_utm,
  shore_utm
)

shore_distance <- st_distance(
  surf_utm,
  shore_utm[nearest_shore, ],
  by_element = TRUE
)

surf_utm$swim_distance_km <-
  as.numeric(shore_distance) / 1000

# Illustrative swimming speed
swim_speed_kmh <- 2

surf_utm$swim_time_min <-
  surf_utm$swim_distance_km /
  swim_speed_kmh * 60

surf_utm$swim_category <- cut(
  surf_utm$swim_time_min,
  breaks = c(-Inf, 5, 10, 20, Inf),
  labels = c(
    "0-5 min",
    "5-10 min",
    "10-20 min",
    "20+ min"
  )
)
```

```{r}
#| label: swimming-table

swim_results <- surf_utm |>
  st_drop_geometry() |>
  select(
    swim_distance_km,
    swim_time_min,
    swim_category
  )

knitr::kable(
  head(swim_results, 15),
  digits = 2,
  caption = "Illustrative shoreline-to-site accessibility"
)
```

```{r}
#| label: swimming-map

swim_pal <- colorFactor(
  palette = time_colors,
  domain = levels(surf_utm$swim_category)
)

swim_map <- leaflet() |>
  addProviderTiles("CartoDB.Positron") |>
  addCircleMarkers(
    data = st_transform(surf_utm, 4326),
    radius = 7,
    color = ~swim_pal(swim_category),
    fillColor = ~swim_pal(swim_category),
    fillOpacity = 0.85,
    popup = ~paste0(
      "Shore access distance: ",
      round(swim_distance_km, 2),
      " km<br>",
      "Hypothetical swim time: ",
      round(swim_time_min, 1),
      " min"
    )
  ) |>
  addCircleMarkers(
    data = shoreline,
    radius = 4,
    color = "orange",
    group = "Shoreline Access"
  ) |>
  addLegend(
    position = "bottomright",
    pal = swim_pal,
    values = surf_utm$swim_category,
    title = "Hypothetical Swim Time"
  ) |>
  setView(
    lng = -157.95,
    lat = 21.48,
    zoom = 10
  )

swim_map
```

# Coast Guard Response Accessibility

```{r}
#| label: coast-guard-table

if (nrow(coast_guard) > 0) {
  knitr::kable(
    coast_guard,
    caption = "Available Coast Guard facility records"
  )
} else {
  cat(
    "Coast Guard facility data has not been added."
  )
}
```

```{r}
#| label: coast-guard-map

has_coordinates <- all(
  c("longitude", "latitude") %in% names(coast_guard)
)

if (has_coordinates) {

  valid_rows <- !is.na(coast_guard$longitude) &
    !is.na(coast_guard$latitude)

  valid_facilities <- coast_guard[
    valid_rows,
    ,
    drop = FALSE
  ]

} else {

  valid_facilities <- coast_guard[FALSE, , drop = FALSE]

}

if (nrow(valid_facilities) > 0) {

  coast_guard_sf <- st_as_sf(
    valid_facilities,
    coords = c("longitude", "latitude"),
    crs = 4326
  )

  leaflet(coast_guard_sf) |>
    addProviderTiles("CartoDB.Positron") |>
    addCircleMarkers(
      radius = 8,
      color = "purple",
      popup = "Coast Guard facility"
    ) |>
    setView(
      lng = -157.95,
      lat = 21.48,
      zoom = 10
    )

} else {

  cat(
    "Coast Guard facility coordinates are not yet ",
    "verified. No station point is plotted."
  )

}
```

```{r}
#| label: marine-time-function

marine_time <- function(
  distance_nm,
  speed_knots,
  mobilization_min = 0
) {

  if (
    any(distance_nm < 0) ||
    any(speed_knots <= 0) ||
    any(mobilization_min < 0)
  ) {
    stop("Invalid marine travel-time inputs.")
  }

  mobilization_min +
    (distance_nm / speed_knots) * 60
}

# Mathematical example only:
# 10 nautical miles at 20 knots.
marine_time(
  distance_nm = 10,
  speed_knots = 20
)
```

```{r}
#| label: comparison-table

comparison <- data.frame(
  Service = c(
    "Lifeguard Proximity",
    "Swimming / Shoreline Access"
  ),
  Mean_Distance_km = c(
    mean(surf_utm$lifeguard_distance_km),
    mean(surf_utm$swim_distance_km)
  ),
  Mean_Time_min = c(
    mean(surf_utm$lifeguard_time_min),
    mean(surf_utm$swim_time_min)
  ),
  Maximum_Time_min = c(
    max(surf_utm$lifeguard_time_min),
    max(surf_utm$swim_time_min)
  )
)

knitr::kable(
  comparison,
  digits = 2,
  caption = "Preliminary hypothetical accessibility scenarios"
)
```

```{r}
#| label: comparison-chart

ggplot(
  comparison,
  aes(
    x = Service,
    y = Mean_Time_min,
    fill = Service
  )
) +
  geom_col(show.legend = FALSE) +
  labs(
    title = "Mean Hypothetical Travel Time",
    x = "Accessibility Scenario",
    y = "Minutes"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(
      angle = 15,
      hjust = 1
    )
  )
```

```{r}
#| label: export-results

dir.create(
  "outputs",
  showWarnings = FALSE
)

st_write(
  st_transform(surf_utm, 4326),
  "outputs/ocean_rescue_accessibility.geojson",
  delete_dsn = TRUE,
  quiet = TRUE
)

write.csv(
  comparison,
  "outputs/ocean_rescue_summary.csv",
  row.names = FALSE
)

cat("Results saved in the outputs folder.")
```

# Limitations

This analysis is preliminary and exploratory.

The lifeguard and swimming accessibility models use straight-line distances and hypothetical travel speeds.

The models do not account for actual dispatch, staff availability, currents, waves, reefs, equipment, or rescue procedures.

Body-surfing sites are used as sample coastal destinations rather than observed emergency incident locations.

The study area is defined using an approximate bounding box rather than an official island boundary.

Coast Guard response times are not calculated because operational marine departure locations and navigable routes have not been verified.

# Conclusion

This project provides an initial geospatial assessment of swimming and lifeguard accessibility across Oʻahu.

It also establishes a data framework for future Coast Guard marine accessibility analysis.

The preliminary models help demonstrate how geographic distance and assumed travel speeds can be used to investigate potential coastal accessibility differences.

Further operational and marine data are required before these results can support conclusions about actual emergency response performance.
