---
title: "Isochrones"
format: html
editor: visual
---

## What is an Isochrone?

An **isochrone** is a line or polygon on a map that connects all locations that can be reached from a specific starting point within a given amount of time.

For this analysis, I will create a drive-time accessibility surface centered on **Chaminade University of Honolulu**.

[Inspiration from Kyle Walker](https://walker-data.com/posts/accessibility-surface/)

## Libraries

```{r}
library(mapboxapi)
library(leaflet)
library(mapview)
library(sf)
library(fasterize)
library(raster)
```

## Create Isochrones

First, create drive-time isochrones around Chaminade University.

```{r}
isos <- mb_isochrone(
  location = "3140 Waialae Ave, Honolulu, HI 96816",
  profile = "driving",
  time = 1:45
)
```

The `time = 1:45` argument creates drive-time boundaries for each minute from 1 to 45 minutes.

## View the Isochrones

Before creating the accessibility surface, we can inspect the individual isochrone polygons.

```{r}
mapview(
  isos,
  zcol = "time"
)
```

## Create an Accessibility Surface

The isochrone polygons can be converted into a raster surface that represents estimated drive time from Chaminade University.

### Reproject the Isochrones

Because raster calculations work better in a projected coordinate system, transform the data to **UTM Zone 4N**, which covers Honolulu.

```{r}
isos_proj <- st_transform(
  isos,
  crs = 32604
)
```

### Create a Raster Template

Create a raster grid with a resolution of 100 meters.

```{r}
template <- raster(
  isos_proj,
  resolution = 100
)
```

### Rasterize the Isochrones

Convert the isochrone polygons into a raster surface using drive time as the raster value.

```{r}
iso_surface <- fasterize(
  isos_proj,
  template,
  field = "time",
  fun = "min"
)

iso_surface
```

## Create the Color Palette

Create a continuous color scale based on drive time.

```{r}
pal <- colorNumeric(
  palette = "plasma",
  domain = isos$time,
  na.color = "transparent"
)
```

## Create the Chaminade Logo Marker

Create a custom Leaflet icon using the Chaminade athletics logo.

```{r}
chaminade_icon <- makeIcon(
  iconUrl = "images/Athletics_C.png",
  iconWidth = 50,
  iconHeight = 50
)
```

## Get Chaminade's Coordinates

Use Mapbox geocoding to retrieve the longitude and latitude of Chaminade University.

```{r}
chaminade <- mb_geocode(
  "3140 Waialae Ave, Honolulu, HI 96816"
)

chaminade
```

## Create the Final Accessibility Map

Now combine the basemap, accessibility surface, legend, and Chaminade logo into one interactive Leaflet map.

```{r}
chaminade_map <- leaflet() %>%
  
  addMapboxTiles(
    style_id = "light-v9",
    username = "mapbox",
    scaling_factor = "0.5x"
  ) %>%
  
  addRasterImage(
    iso_surface,
    colors = pal,
    opacity = 0.5
  ) %>%
  
  addLegend(
    values = isos$time,
    pal = pal,
    title = "Drive-time from<br>Chaminade University"
  ) %>%
  
  addMarkers(
    lng = chaminade[1],
    lat = chaminade[2],
    icon = chaminade_icon,
    popup = "Chaminade University of Honolulu"
  )

chaminade_map
```

## Going Further

The Chaminade accessibility surface demonstrates how isochrones can be used to visualize travel time from a single location.

A larger project could use the same method to investigate access to important emergency services across Oʻahu.

### Research Question

**How accessible are critical emergency services across Oʻahu based on driving time?**

Potential locations could include:

- fire_stations: mb_isochrone works well

  -ambulance:

  -shelters:

  can't go off based travel time in island wide emergency situations

  -hospital, trauma centers, etc,.

  -police stations, hostage situation

  -swim, lifeguard response, coast guard response -DJ

  The locations could eventually be combined into an interactive **Shiny application** where users could select different types of emergency services and compare travel-time accessibility across the island.

```{r}
library(sf)
library(dplyr)
library(leaflet)
library(mapview)
library(ggplot2)
library(units)
library(osmdata)
```
